Parsing Postfix log files

### 2009-04-11 doc/logparser-results.tex
Corrections/improvements after proofreading second coverage section.

Full coverage of log lines is easy to achieve, yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files; \parsername{}
has \numberOFrules{} rules, fully parsing \numberOFlogFILESall{} log files
from Postfix 2.2 through to Postfix 2.5.  Maintaining full coverage is hard
because log lines change over time, e.g.\ administrators add restrictions
with custom messages, \acronym{DNSBL} messages change, or major releases of
Postfix change log lines (usually adding more information).  Warnings are
issued for any log lines that are not recognised; no warnings are issued
for unrecognised log lines while parsing the \numberOFlogFILES{} log files,
so it can be safely concluded that there are zero false negatives.  False
positives are harder to quantify: short of examining each of the 60,721,709
log lines and determining which rule recognised it, there is no way to be
sure that every log line was recognised by the correct rule, making it
impossible to quantify the false positive rate.  However, a random sample
of 6,039 log lines was parsed, and the results manually verified by
inspection to ensure that the correct rule recognised each log line.  The
sample was generated by running the command: \texttt{perl -n -e
\singlequote{}print if (rand 1 < 0.0001)\singlequote{} LOG\_FILES} to
randomly extract roughly one log line in every 10,000 (it actually
extracted 0.00994\% instead of 0.001\%).  Each log line was examined and
the correct rule identified from the \numberOFrules{} rules in the
database; the correct rule was then compared to the rule that recognised
the log line when parsing.  The sample results contained zero false
positives, and this check has been automated to ensure continued accuracy.
Based on this, the author is confident that zero false positives occur when
parsing the \numberOFlogFILES{} log files.  On initial appearances,
exercising only 36 rules from a total of \numberOFrules{} when parsing
6,039 log lines seems low, but after examining \graphref{rule hits graph}
it becomes apparent that parsing using such a low number of rules is to be
expected.  The reader should also bear in mind that even when parsing all
\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} are used,
because some of the rules are for recognising log lines that only appear in
other log files.

\subsection{Mails Covered}

@@

Coverage of mails is much more difficult to determine accurately than
coverage of log lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails that were not handled properly (many of the
complications discussed in \sectionref{complications} were detected in this
way).  The parser issues warnings when it detects any errors, some of which
may alert the user to a problem, e.g.\ when a queueid is reused before the
previous mail is fully dealt with, when a queueid or \acronym{pid} is not
found in the state tables,\footnote{There will often be warnings about a
missing queueid or \acronym{pid} in the first few hundred or thousand log
lines because the earlier log lines for those connections or mails are in
the previous log file; loading the saved state from the previous log file
will solve this problem.} or when there are problems tracking a child mail
(see \sectionref{tracking re-injected mail}).  There should be few or no
warnings when parsing, and when finished parsing the state table should
only contain entries for mails that had yet to be delivered when the log
files ended, or were accepted before the log files began.

At the time of writing the parser is being tested with \numberOFlogFILES{}
log files.  There are 5 warnings produced, but because the parser errs on
the side of producing more warnings rather than fewer, those 5 warnings
represent 3 instances of 1 problem: 3 connections started before the first
log file, so their initial log lines are missing, leading to warnings when
their remaining log lines are parsed.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails that
are awaiting delivery after the period covered by the log files, though
they may also contain mails whose initial entries are not contained in the
log files.  Any other entries are evidence of a failure in parsing or an
aberration in the log files.  After parsing the \numberOFlogFILES{} test
log files the state tables contain 18 entries, breaking down into:

@@

    \item 9 mails whose initial log lines were not present in the log
        files.

@@

There are no mails in the state tables which should not be present, thus it
can be concluded that there are zero false negatives.  Once again,
determining the false positive rate is much harder, as manually checking
the results of parsing 13,850,793 connections and mails accepted, rejected,
bounced or delivered is infeasible.  There is considerable circumstantial
evidence that the false positive rate is quite low:

@@

    \item The parser is quite verbose when complaining about known problems
        (e.g.\ if a mail is tracked twice as described in
        \sectionref{tracking re-injected mail}), and no such warnings are
        produced during the test runs.

    \item Queueids and \glspl{pid} naturally group together log lines
        belonging to one mail or connection respectively; it is extremely
        unlikely that a log line would be associated with the wrong
        connection.

    \item When dealing with the complications described in
        \sectionref{complications} the solutions are as specific and
        restrictive as possible, with the goal of minimising the number of
        false positives.  In addition the solution to the \textit{Out of
        order log files\/} complication described in \sectionref{out of
        order log lines} imposes conditions that each reassembled mail must
        comply with to be acceptable.

    \item Every effort has been made while developing to make the parser as
        precise, demanding, and particular as possible.

@@

Whereas verifying by inspection that the parser correctly deals with all
60,721,709 log lines in the test log files is infeasible, verifying a
subset of those log files is a tractable, if extremely time consuming,
task.  A sample of log lines was obtained by randomly selecting a log file:

\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline{}
\verb!            -e 'print [shuffle(@ARGV)]->[0];'!

The first 6000 log lines of this log file (roughly 0.01\% of the total
number of log lines used in testing) was extracted:

\verb!    sed -n -e '1,6000p' logfile > test-log-segment!

It is important that the log lines used are contiguous so that all log
entries are present for as many of the connections as possible.  This log
segment was parsed with all debugging options enabled, resulting in 167,448
lines of output.\footnote{A mean of 27.908 lines of output per line of
input; each connection has 30 debugging lines, plus 21 debugging lines per
result.  Connections which have been cloned will have the cloned connection
in their debugging output, plus another 33 debugging lines.  Those numbers
are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
between the number of log lines and debugging lines: $33(connections) +
30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
and has not been rigorously verified.}  All 167,448 lines were examined in
conjunction with the input log file and a dump of the resulting database,
verifying that for each of the log lines the parser used the correct rule
and executed the correct action, which in turn produced the correct result
and inserted the correct data in the database.  The log file segment
produced 4 warnings, 10 mails remaining in the state tables, and 1625
connections correctly entered in the database.

Given the circumstantial and experimental evidence detailed above, the
author is confident that the false positive rate when reconstructing a mail
is exceedingly low, if not approaching zero.

@@

>>>>

Full coverage of log lines is easy to achieve, yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files; \parsername{}
has \numberOFrules{} rules, fully parsing \numberOFlogFILESall{} log files
from Postfix 2.2 through to Postfix 2.5.  Maintaining full coverage is hard
because log lines change over time, e.g.\ administrators add restrictions
with custom messages, \acronym{DNSBL} messages change, or major releases of
Postfix change log lines (usually adding more information).  Warnings are
issued for any log lines that are not recognised; no warnings are issued
for unrecognised log lines while parsing the \numberOFlogFILES{} log files,
so it can be safely concluded that there are zero false negatives.  False
positives are harder to quantify: short of examining each of the 60,721,709
log lines and determining which rule recognised it, there is no way to be
sure that every log line was recognised by the correct rule, making it
impossible to quantify the false positive rate.  However, a random sample
of 6039 log lines was parsed, and the results manually verified by
inspection to ensure that the correct rule recognised each log line.  The
sample was generated by running the command: \texttt{perl -n -e
\singlequote{}print if (rand 1 < 0.0001)\singlequote{} LOG\_FILES} to
randomly extract roughly one log line in every 10,000 (it actually
extracted 0.00994\% instead of 0.01\%).  Each log line was examined and the
correct rule identified from the \numberOFrules{} rules in the database;
the correct rule was then compared to the rule that recognised the log line
when parsing.  The sample results contained zero false positives, and this
check has been automated to ensure continued accuracy.  Based on this, the
author is confident that zero false positives occur when parsing the
\numberOFlogFILES{} log files.  On initial appearances, exercising only 36
rules from a total of \numberOFrules{} when parsing 6039 log lines seems
low, but after examining \graphref{rule hits graph} it becomes apparent
that parsing using such a low number of rules is to be expected.  The
reader should also bear in mind that even when parsing all
\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} are used,
because some of the rules are for recognising log lines that only appear in
other log files.

\subsection{Mail Delivery Attempts Correctly Understood And Reconstructed}

@@

The proportion of mail delivery attempts that are correctly understood and
reconstructed is much more difficult to determine accurately than the
proportion of log lines that are correctly recognised.  The parser can dump
its state tables in a human readable form; examining those tables with
reference to the log files is the best way to detect mails that were not
handled properly (many of the complications discussed in
\sectionref{complications} were detected in this way).  \parsername{}
issues warnings when it detects any errors or discrepancies, alerting the
user to the problem, e.g.\ when a queueid is reused but the previous mail
remains in the state tables, when a queueid or \acronym{pid} is not found
in the state tables, or when a mail does not include sufficient data to
satisfy the database schema.  There should be few or no warnings during
parsing, and when finished parsing the state tables should only contain
entries for mails that have log lines in subsequent log files.  There will
often be warnings about a missing queueid or \acronym{pid} in the first few
thousand log lines, because the earlier log lines for those connections or
mails are in a previous log file; loading saved state tables from the
previous log file will solve this problem.

The data used in this chapter is generated by parsing \numberOFlogFILES{}
log files.  There are 5 warnings produced, but because \parsername{} errs
on the side of producing more warnings rather than fewer, those 5 warnings
represent 3 instances of 1 problem: 3 connections that started before the
first log file, so their initial log lines are missing, leading to warnings
when their remaining log lines are parsed.  There are no false positives.

The state tables will contain entries for mails not yet delivered when the
parser finishes execution.  Ideally, that is all they will contain, though
they may also contain mails whose initial log lines are not contained in
the log files.  Any other entries in the state tables are evidence of
either a failure in parsing, or an aberration in the log files.  After
parsing the \numberOFlogFILES{} log files, the state tables contain 18
entries, breaking down into:

@@

    \item 9 mails whose initial log lines were not present in the log
        files.  XXX WHY WERE THERE NO WARNINGS ABOUT THESE\@?

@@

There are no mails in the state tables which should not be present, thus it
can be concluded that there are zero false negatives.  Once again,
determining the false positive rate is much harder, as manually checking
the results of parsing 13,850,793 connections and mails accepted, rejected,
bounced, or delivered is infeasible.  There is considerable evidence that
the false positive rate is extremely low:

@@

    \item The parser is quite verbose when complaining about known
        problems, e.g.\ if a mail is missing required data, and no such
        warnings are produced during the test runs.

    \item Queueids and \glspl{pid} naturally identify log lines belonging
        to one mail or connection respectively; it is extremely unlikely
        that a log line would be associated with the wrong connection.

    \item When dealing with the complications described in
        \sectionref{complications}, the solutions are as specific and
        restrictive as possible, with the goal of minimising the number of
        false positives.  In addition, the solution to the complication
        described in \sectionref{out of order log lines} imposes conditions
        that each reconstructed mail must comply with to be acceptable.

    \item Every effort has been made to make \parsername{} as precise,
        demanding, and particular as possible.

@@

%\verb!perl -Mstrict -Mwarnings -MList::Util=shuffle -e! \newline{}
%\verb!     '@ARGV = [shuffle(@ARGV)]->[0];!             \newline{}
%\verb!      while (<>) {!                               \newline{}
%\verb!          if (rand 1 < 0.0001) {!                 \newline{}
%\verb!              foreach my $i (1 .!.  6000) {!        \newline{}
%\verb!                  print scalar <>;!               \newline{}
%\verb!              }!                                  \newline{}
%\verb!              exit;!                              \newline{}
%\verb!          }!                                      \newline{}
%\verb!      }' LOG_FILES!

Verifying by inspection that the parser correctly deals with all 60,721,709
log lines in the \numberOFlogFILES{} log files is infeasible, but verifying
the parsing of a sample from those log lines is a tractable, if extremely
time consuming, task.  A sample of log lines was obtained by randomly
selecting a log file, and then randomly selecting a block of 6000
contiguous log lines from it (0.00988\% of the total number of log lines).
It is important that the log lines are contiguous, so that all log lines
are present for as many of the mail delivery attempts contained in the
block as possible.  This log segment was parsed with all debugging options
enabled, resulting in 167,448 lines of output.\footnote{A mean of 27.908
lines of output per line of input; each connection has 30 debugging lines,
plus 21 debugging lines per result.  Connections which have been cloned
will have the cloned connection in their debugging output, plus another 33
debugging lines.  Those numbers are approximate, and may vary $\pm{}$ 2.
There is a linear relationship between the number of log lines and
debugging lines: $33(connections) + 30(accepted~~mails) + 21(results)$.
This formula is an approximation only, and has not been rigorously
verified.}  All 167,448 lines were examined in conjunction with the log
file segment and a dump of the resulting database, verifying that for each
of the log lines \parsername{} used the correct rule and invoked the
correct action, which in turn produced the correct result, and the correct
data was inserted in the database.  The log file segment produced 4
warnings, 10 mails remaining in the state tables, 1625 mail delivery
attempts correctly entered in the database, zero false positives, and zero
false negatives.

Given the evidence detailed above, the author is confident that the false
positive rate when reconstructing a mail delivery attempt from the
\numberOFlogFILES{} log files is exceedingly low, if not zero.

@@

Experience implementing the Postfix log parser shows that full input
coverage is relatively easy to achieve with this architecture, and that
with enough time and effort full understanding of the input is possible.
Postfix log files would require substantial time and effort to correctly
parse regardless of the architecture used; this architecture enables an
iterative approach to be used (similar to Stepwise
Refinement~\cite{stepwise-refinement}), as is practiced in many other
software engineering disciplines.
